Scaling rates of true polar wander in convecting planets and moons

### 2016-07-18 tpw_rate.tex
Small fixes.
Light edits to help with parsing.

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}

@@

\linenumbers

@@

The primary uncertainties in assigning a maximum TPW rate to a convecting planet are the size of convective anomalies, which drive the rotational adjustment, and the viscosity structure of the mantle, which retards it. 
These two uncertainties are not unrelated: they are both expected to be functions of the geometric and material properties of the mantle.
As such, they do not vary independently, and first-order questions about the propensity for planets to experience TPW remain: how are rates of TPW expected to vary with the vigor of convection? 
Are other planetary bodies more or less likely than Earth to experience TPW?  And are these rates expected to vary in Earth history?

@@

\begin{equation}
\frac{d \mathbf{H}}{dt} + \mitbf{\Omega} \times \mathbf{H} = 0
\label{eq:euler}
\end{equation}
where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
In this case Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
For a slowly convecting fluid such as a planetary mantle the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
\begin{equation}
\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0.
\label{eq:quasistatic_liouville}
\end{equation}
Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for $\mathbf{I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
The moment of inertia tensor in Equation~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
Here we are interested in contributions from mantle convection, so we restrict our attention to these processes.

@@

\subsection{A note on reference frames}
\label{sec:reference_frames}
True polar wander can be described in different reference frames, and this choice is fundamentally an arbitrary one.
However, certain aspects of the physics can be made much simpler by an appropriate choice of the reference frame.
In our treatment of TPW, we will refer to three different reference frames:

First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
Second, there is the body-fixed (or geographic) frame.
By definition, the rotation of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
A terrestrial no-net-rotation or hotspot reference frame are common choices for the body-fixed frame.
Treatments of gravitational or rotational deformation of a planet are most naturally expressed in the body-fixed frame,
as described in Section~\ref{sec:rotational_deformation}.
Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
which we will refer to as the ``$\mathbf{E}$-frame.'' 
Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
causing the $\mathbf{E}$-frame to slowly rotate with respect to the geographic frame. We denote this drift 
by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
This, as well, may be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
and $C_{ij}$ is the moment of inertia due solely to the internal load.
Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
permits more general treatments of mantle rheology.

@@

As a warmup, consider the case where $\phi = 0$, then  
\begin{equation}
\mitbf{\omega} = 
\begin{bmatrix}
\sin{\theta}  \\
0 \\
\cos{\theta} \\
\end{bmatrix}
\end{equation}
We can evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \\
0 \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

@@

Now consider the more general case whhere $\phi \ne 0$.
Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \cos{\phi}\\
\lambda_2 \sin{\theta} \sin{\phi} \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

@@

The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_3$ plane ($\phi = 0$).
Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
\begin{equation}
\vert \dot{\Theta} \vert = 
\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1).
\label{eq:simple_milankovitch}
\end{equation}
The maximum polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).

From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating $\dot{\Theta}$ are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.  
They represent, respectively, the size of convective anomalies in the moment of inertia tensor 
and the angular mismatch between the rotation axis and the principal axis of the convective moment. 
The dynamics of mantle convection affects the rate of TPW in two ways; 
it controls $\lambda_i$ and it represents the mechanism that drives the angular mismatch. 
The value of $\theta$ at any time also depends on the relaxation of the rotation axis back toward the principal axis of $\mathbf{E}$. 
We focus on the dynamics of the convective contribution in the next section.

@@

Mantle convection and rotational dynamics of planetary bodies are usually considered separately, yet the processes are based on a common set of governing equations. 
As such, some extra care must be taken to ensure that the equations we consider are self-consistent. 
Furthermore, since our goal is to establish a scaling for TPW rate, we must identify a minimal set of nondimensional numbers which describe its physics.
For simplicty we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.  The equations for mass, momentum, and energy then read

@@

\begin{equation}
\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
\label{eq:energy}
\end{equation}
where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
The vector $\mathbf{g}$ is the gravitational acceleration, 
which defined in terms of a gravitational potential $V$ by $\mathbf{g} = -\nabla V$. 
The gravitational potential obeys 
\begin{equation}
\nabla^2 V = 4 \pi G \rho.
\end{equation}
where $G$ is the gravitational constant. 
For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is the approximately constant (as is the case for Earth's mantle).
In addition we use the simple equation of state
\begin{equation}
\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right).
\label{eq:eos}
\end{equation}
to define the density.
The remaining parameters are defined in Table~\ref{tab:parameters}.
Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure
(in the latter case the boundary conditions on $P$ must be modified).
The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
is generally small compared to gravitational forces (at least for Earthlike parameters),
and likely does not have a strong influence on the style of convection.
However, it is critical for determining the the size of the rotational bulge which must be moved around via TPW, 
and so we must include it to establish the connection between the linear and angular momentum equations.

@@

In this case the dynamics can be characterized in terms of deviations from a reference hydrostatic state, 
which includes the dynamic pressure $P^* = P - P_0$ and density perturbations $\delta \rho = \rho- \rho_0 = - \rho_0 \alpha (T-T_0)$.
In addition, we expect deviations in the figure of the planet from its hydrostatic shape, 
which we denote by $V = V_H + \Delta V$, where $V_H$ is the hydrostatic figure and $\Delta V$ is the deviation. 
Our rationale for this decomposition is simply that the hydrostatic pressure is most naturally defined in the (oblate) hydrostatic configuration. 
The introduction of $\Delta V$ requires another nondimensional number to characterize it, 
and we find that the quantity $\Gamma \equiv \alpha \Delta T$ is convenient.
Finally, we define $\mitbf{\Omega} = \Omega_0 \mitbf{\omega}$, where $\mitbf{\omega}$ is a unit vector in the direction of $\mitbf{\Omega}$.

@@

\begin{equation}
\begin{aligned}
\int_V \rho \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV &= 0 \\
\label{eq:quasistatic_angular_momentum}
\end{aligned}
\end{equation}
which may be rewritten via the Jacobi identity to find
\ifdetail
\begin{equation}
\begin{aligned}
- \int_V \rho (\mitbf{\Omega}\times \mathbf{r}) \times ( \mathbf{r} \times \mitbf{\Omega} ) \,dV + 
- \int_V \rho \mitbf{\Omega}\times(\mitbf{\Omega}\times \mathbf{r}) \times \mathbf{r} \,dV &= 0 \\
\end{aligned}
\end{equation}
\fi
\begin{equation}
\begin{aligned}
 \mitbf{\Omega} \times \int_V \rho \mathbf{r} \times (\mitbf{\Omega}\times \mathbf{r}) \,dV &= 0. \\
\end{aligned}
\end{equation}
This can be directly identified with Equation~\eqref{eq:quasistatic_liouville},
and is a statement that a quasistatic body will rotate around the principal axis of its total moment of inertia.

@@

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is 
proportional to the ratio of rotational to gravitational forces \citep{munk1960rotation}:
\begin{equation}
(C-A) \sim I_0 m.
\end{equation}

@@

The quantity $\Lambda_{ij}$ should be a function of our nondimensional parameters $\mathrm{Ra}$, $m$, and $\Gamma$. 
Similarly, the convective forcing that drives changes in $\theta$ must also be a function these nondimensional parameters. 
To be explicit in our discussion, we denote this driving force by ${\Psi}$.
We consider relatively slowly rotating bodies here ($m \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
In this regime, therefore, we neglect the dependence on $m$, 
and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra}, \Gamma)$ and $\Psi(\mathrm{Ra}, \Gamma)$.

@@

Fluctuations in the nonhydrostatic moment of inertia are due to temperature fluctuations in the mantle and their spatial structure:

\begin{equation}
C_{ij} = \int_{V_H} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) \,dV
\label{eq:temperature_fluctuations}
\end{equation}
As discussed in Section~\ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
The surface response is a function of the viscosity structure of the planet and the wavelength and depth of the interal load.
The factor $(1+k^L_f)$ is generally an order-one parameter 
(strictly speaking, dynamic compensation usually makes it less than one, \citep[e.g.][]{richards1984geoid}). 
For the purposes of scaling it is reasonable to neglect this multiplicative factor.

For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
Therefore, an estimate for one specifies the other.
We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
$T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$:
\begin{equation} 
T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
\label{eq:T_series}
\end{equation}
Inserting the temperature expansion~\eqref{eq:T_series} into Equation~\eqref{eq:temperature_fluctuations} results in 
a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
Orthogonality of the basis functions for the expansion means that the integral for $C_{ij}$ picks out degree-two 
spherical harmonics in the lateral dimensions, and only the lowest few radial functions $R_n(r)$.
Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix~\ref{appendix:moments} for more detail): 
\begin{equation}
\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
\end{equation}
The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by

\begin{equation}
\max \left( \displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \right) = 1.
\end{equation}
This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.  
We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow (relatively low $\mathrm{Ra}$) 
and that of chaotic flow (relatively high $\mathrm{Ra}$).
The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is a broadening of the spatial 
and temporal spectra \citep{mclaughlin1982transition}.

@@

With an estimate for the power in the degree-two part of the temperature field, we may finally estimate $\Lambda_{ij}$:
\begin{equation}
\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }.
\label{eq:lambda_estimate}
\end{equation}
Other power spectra for the temperature field are possible. Isoviscous models tend to be ``bluer,'' and models with viscosity stratification tend to be ``redder'' \citep{richards1999polar}.
Present day Earth seems to have a fairly ``red'' spectrum, with large low-degree seismic anomalies due to 
Cenozoic subduction history and the lower mantle LLSVPs \citep{dziewonski2010mantle}.
Nevertheless, at high Rayleigh number the expectation is that the power will be distributed across many length scales.

\subsection{An estimate for the angular mismatch $\theta$}
\label{sec:theta}

Convection drives a mismatch $\theta$ between the current rotation axis and the principal axis of the convective moment. 
This parameter is crucial for the rate of TPW, as previously noted in Equation~\eqref{eq:scaled_rotation}.  
Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.

@@

In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
the mismatch angle is $90^\circ$. However, the eigengap does not need to be zero for there to be large displacements polar wander, 
and if it does go to zero, the wander does not need to be $90^\circ$.

>>>>

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%%\usepackage{lineno}

@@

%%\linenumbers

%%BEGINCLIP

@@

The primary uncertainties in assigning a maximum TPW rate to a convecting planet are the size of convective anomalies, which drive the rotational adjustment, and the viscosity structure of the mantle, which retards it. 
These two uncertainties are not unrelated: they are both expected to be functions of the geometric and material properties of the mantle.
As such, they do not vary independently, and first-order questions about the propensity for planets to experience TPW remain: how are rates of TPW expected to vary with the vigor of convection? 
Are other planetary bodies more or less likely than Earth to experience TPW?  And are these rates expected to vary through Earth history?

@@

\begin{equation}
\frac{d \mathbf{H}}{dt} + \mitbf{\Omega} \times \mathbf{H} = 0
\label{eq:euler}
\end{equation}
where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
In this case, Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
For a slowly convecting fluid, such as a planetary mantle, the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
\begin{equation}
\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0.
\label{eq:quasistatic_liouville}
\end{equation}
Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for $\mathbf{I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
The moment of inertia tensor in Equation~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
Here we are interested in contributions from mantle convection, so we restrict our attention to these processes.

@@

\subsection{A note on reference frames}
\label{sec:reference_frames}
True polar wander can be described in different reference frames, and this choice is fundamentally an arbitrary one.
However, certain aspects of the physics can be made much simpler by an appropriate choice of the reference frame.
In our treatment of TPW, we will refer to three different reference frames:
\begin{itemize}
\item First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
\item Second, there is the body-fixed (or geographic) frame.
By definition, the rotation of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
A terrestrial no-net-rotation or hotspot reference frame are common choices for the body-fixed frame.
Treatments of gravitational or rotational deformation of a planet are most naturally expressed in the body-fixed frame,
as described in Section~\ref{sec:rotational_deformation}.
\item Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
which we will refer to as the ``$\mathbf{E}$-frame.'' 
Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
causing the $\mathbf{E}$-frame to slowly rotate with respect to the geographic frame. We denote this drift 
by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).
\end{itemize}

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
This part of the moment of inertia may also be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
and $C_{ij}$ is the moment of inertia due solely to the internal load.
Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
permits more general treatments of mantle rheology.

@@

As a warm-up, consider the case where $\phi = 0$, then  
\begin{equation}
\mitbf{\omega} = 
\begin{bmatrix}
\sin{\theta}  \\
0 \\
\cos{\theta} \\
\end{bmatrix}
\end{equation}
We can evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \\
0 \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

@@

Now consider the more general case where $\phi \ne 0$.
Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \cos{\phi}\\
\lambda_2 \sin{\theta} \sin{\phi} \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

@@

The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_3$ plane ($\phi = 0$).
Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
\begin{equation}
\vert \dot{\Theta} \vert = 
\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1).
\label{eq:simple_milankovitch}
\end{equation}
The maximum true polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).

From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating the rate of true polar wander ($\dot{\Theta}$)
are $\theta$ and the differences between the eigenvalues of the convective moment $\mathbf{E}$, both of which depend on the structure and dynamics of mantle convection.  
They represent, respectively, the angular mismatch between the rotation axis and the principal axis of the convective moment 
and the size of convective anomalies in the moment of inertia tensor. 
The dynamics of mantle convection affects the rate of TPW in two ways; 
it controls $\lambda_i$ and it represents the mechanism that drives the angular mismatch. 
The value of $\theta$ at any time also depends on the relaxation of the rotation axis back toward the principal axis of $\mathbf{E}$. 
We focus on the dynamics of the convective contribution in the next section.

@@

Mantle convection and rotational dynamics of planetary bodies are usually considered separately, yet the processes are based on a common set of governing equations. 
As such, some extra care must be taken to ensure that the equations we consider are self-consistent. 
Furthermore, since our goal is to establish a scaling for TPW rate, we must identify a minimal set of nondimensional numbers which describe its physics.
For simplicity we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.  The equations for mass, momentum, and energy then read

@@

\begin{equation}
\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
\label{eq:energy}
\end{equation}
where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
The vector $\mathbf{g}$ is the gravitational acceleration, 
which defined in terms of a gravitational potential $V$ by $\mathbf{g} = -\nabla V$. 
The gravitational potential obeys 
\begin{equation}
\nabla^2 V = 4 \pi G \rho.
\end{equation}
where $G$ is the gravitational constant. 
For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is the approximately constant (as is the case for Earth's mantle).
In addition we use the simple equation of state
\begin{equation}
\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right).
\label{eq:eos}
\end{equation}
to define the density.
The remaining parameters are defined in Table~\ref{tab:parameters}.
Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure
(in the latter case the boundary conditions on $P$ must be modified).
The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
is generally small compared to gravitational forces (at least for Earth-like parameters),
and likely does not have a strong influence on the style of convection.
However, this term is critical for determining the size of the rotational bulge which must move relative to the mantle,
and so we must include it to establish the connection between the linear and angular momentum equations.

@@

In this case, the dynamics can be characterized in terms of deviations from a reference hydrostatic state, 
which includes the dynamic pressure $P^* = P - P_0$ and density perturbations $\delta \rho = \rho- \rho_0 = - \rho_0 \alpha (T-T_0)$.
In addition, we expect deviations in the figure of the planet from its hydrostatic shape, 
which we denote by $V = V_H + \Delta V$, where $V_H$ is the hydrostatic figure and $\Delta V$ is the deviation. 
Our rationale for this decomposition is simply that the hydrostatic pressure is most naturally defined in the (oblate) hydrostatic configuration. 
The introduction of $\Delta V$ requires another nondimensional number to characterize it, 
and we find that the quantity $\Gamma \equiv \alpha \Delta T$ is convenient.
Finally, we define $\mitbf{\Omega} = \Omega_0 \mitbf{\omega}$, where $\mitbf{\omega}$ is a unit vector in the direction of $\mitbf{\Omega}$.

@@

\begin{equation}
\begin{aligned}
\int_V \rho \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV &= 0 \\
\label{eq:quasistatic_angular_momentum}
\end{aligned}
\end{equation}
which may be rewritten via the Jacobi identity to find
\ifdetail
\begin{equation}
\begin{aligned}
- \int_V \rho (\mitbf{\Omega}\times \mathbf{r}) \times ( \mathbf{r} \times \mitbf{\Omega} ) \,dV + 
- \int_V \rho \mitbf{\Omega}\times(\mitbf{\Omega}\times \mathbf{r}) \times \mathbf{r} \,dV &= 0 \\
\end{aligned}
\end{equation}
\fi
\begin{equation}
\begin{aligned}
 \mitbf{\Omega} \times \int_V \rho \mathbf{r} \times (\mitbf{\Omega}\times \mathbf{r}) \,dV &= 0. \\
\end{aligned}
\end{equation}
This equation can be directly identified with Equation~\eqref{eq:quasistatic_liouville},
and is a statement that a quasistatic body will rotate around the principal axis of its total moment of inertia.

@@

It is useful to define a nondimensional eigenvalue difference, which may also be described as a normalized ``eigengap'': $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is 
proportional to the ratio of rotational to gravitational forces \citep{munk1960rotation}:
\begin{equation}
(C-A) \sim I_0 m.
\end{equation}

@@

The quantity $\Lambda_{ij}$ should be a function of our nondimensional parameters $\mathrm{Ra}$, $m$, and $\Gamma$. 
Similarly, the convective forcing that drives changes in $\theta$ must also be a function these nondimensional parameters. 
To be explicit in our discussion, we denote the angular velocity induced by forcing by ${\Psi}$ (see Figure~\ref{fig:reference_frames}).
We consider relatively slowly rotating bodies here ($m \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
In this regime, therefore, we neglect the dependence on $m$, 
and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra}, \Gamma)$ and $\Psi(\mathrm{Ra}, \Gamma)$.

@@

Fluctuations in the nonhydrostatic moment of inertia are caused by density variations 
due to the temperature fluctuations in the mantle and their spatial structure:

\begin{equation}
C_{ij} = \int_{V_H} \rho_0 \alpha (T-T_0) \left( r_q r_q \delta_{ij} - r_i r_j \right) \,dV
\label{eq:temperature_fluctuations}
\end{equation}
As discussed in Section~\ref{sec:convective_moment}, these internal density loads may be convolved with their surface responses to get the convective moment of inertia $E_{ij}$.
The surface response is a function of the viscosity structure of the planet and the wavelength and depth of the internal load.
The factor $(1+k^L_f)$ is generally an order-one parameter 
(strictly speaking, dynamic compensation usually makes it less than one, \citep[e.g.][]{richards1984geoid}). 
For the purposes of scaling, it is reasonable to neglect this multiplicative factor.

For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
Therefore, an estimate for one specifies the other.
We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
$T_{nlm}$ are the coefficients for the expansion which have been normalized by $\Delta T$:
\begin{equation} 
T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
\label{eq:T_series}
\end{equation}
Inserting the temperature expansion~\eqref{eq:T_series} into Equation~\eqref{eq:temperature_fluctuations} results in 
a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
Orthogonality of the basis functions for the expansion means that the integral for $C_{ij}$ picks out degree-two 
spherical harmonics in the lateral dimensions, and only the lowest few radial functions $R_n(r)$.
Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix~\ref{appendix:moments} for more detail): 
\begin{equation}
\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
\end{equation}
The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by

\begin{equation}
\max \left( \displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \right) = 1.
\end{equation}
This is a strong constraint, but it gives very little information about the distribution of power across the $T_{lmn}$.  
We can, however, think about the power spectrum in two different regimes: that of steady/quasisteady flow, (relatively low $\mathrm{Ra}$) 
and that of chaotic flow (relatively high $\mathrm{Ra}$).
The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is a broadening of the spatial 
and temporal spectra \citep{mclaughlin1982transition}.

@@

With an estimate for the power in the degree-two part of the temperature field, we may finally estimate $\Lambda_{ij}$:
\begin{equation}
\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }.
\label{eq:lambda_estimate}
\end{equation}
Other power spectra for the temperature field are possible. Isoviscous models tend to be ``bluer,'' (dominated by small wavelengths) 
and models with viscosity stratification tend to be ``redder'' (dominated by long wavelengths) \citep{richards1999polar}.
Present day Earth seems to have a fairly ``red'' spectrum, with large low-degree seismic anomalies due to 
Cenozoic subduction history and the lower mantle LLSVPs \citep{dziewonski2010mantle}.
Nevertheless, at high Rayleigh number the expectation is that the power will be distributed across many length scales.

\subsection{An estimate for the angular mismatch angle ($\theta$)}
\label{sec:theta}

Convection can drive both growth and decay in the mismatch angle ($\theta$) between the current rotation axis and the principal axis of the convective moment. 
This parameter is crucial for the rate of TPW, as previously noted in Equation~\eqref{eq:scaled_rotation}.  
Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.

@@

In the limit that the eigengap becomes zero, the rotation of the principle axes can be arbitrary. 
This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
the mismatch angle is $90^\circ$. However, the eigengap does not need to be zero for there to be large displacements polar wander, 
and if it does go to zero, the wander does not need to be $90^\circ$.

Figure~\ref{fig:misfit} shows a representative timeseries for annular convection, where 
we track the eigenvalues of the moment of inertia and integrate Equation~\eqref{eq:liouville_ode} in time.
Since it is a 2D model, the spin axis may be represented by a single angle.
When the eigengap gets small, the misfit angle becomes much larger, and the rate of polar wander becomes much faster.
At $\sim$1 Gyr into the model run the eigengap goes to zero, and the misfit angle goes to approximately $90^\circ$, 
an IITPW event. But there are several other events where the eigengap becomes small, and there are 
still large TPW events associated with them.
For example, at $\sim$0.3 Gyr the eigengap dips, with an associated large TPW event, even though
there is technically no interchange of the axes.
